Differential growth and transcriptomic profile of stem cell-derived midbrain astrocytes

Version 1

Abstract

Abstract

Regional specificity of stem cell-derived astrocytes is believed to be an important prerequisite for their applications in disease modelling and cell-based therapies. The regional identity of these astrocytes is often defined by the positional characteristics of their antecedent, stem cell-derived neural progenitors patterned to a fate of interest, with the assumption that the positional specification is to be preserved by the derived astrocytes. Using a human induced pluripotent stem cell line designed for tracing midbrain floor plate derivatives, here we show that lineage composition of the derived astrocytes is not a faithful recapitulation of the founder progenitor population, as demonstrated by the loss of floor plate differentiated progeny in the final astrocyte products. Using deep single cell RNA sequencing, we identified distinct transcriptomic signatures of midbrain floor plate-derived astrocytes. Our study highlights the need for rigorous characterisation of pluripotent stem cell-derived regional astrocytes and provides a valuable resource for assessing midbrain floor plate-derived human astrocytes.

Introduction

Astrocytes are the most abundant cells in the brain. They play important roles in the central nervous system in supporting neuronal survival and synaptic activities, including the regulation of ionic homeostasis, providing energetic support, elimination of oxidative stress, and neurotransmitter removal and recycle (Verkhratsky and Nedergaard 2018). Abnormalities in astrocytes have been linked to various neurodegenerative and neurodevelopmental disorders, such as Parkinson’s disease, Alzheimer’s disease, Huntington’s disease, autism spectrum disorders and Alexander’s disease (Molofsky et al. 2012; Phatnani and Maniatis 2015; Booth et al. 2017). There is therefore a growing interest in using human pluripotent stem cell (PSC)-derived astrocytes for disease modelling in vitro (Chandrasekaran et al. 2016).

Contrary to the widely held belief that astrocytes in the brain are largely identical, recent studies revealed diversity in their transcriptomic profile, physiological properties and function (Oberheim et al. 2009; Schober et al. 2022). For example, single cell and spatial transcriptomic studies have identified several astrocyte subpopulations in the mouse cortex (Zhu et al. 2018; Batiuk et al. 2020; Bayraktar et al. 2020). In human, although astrocyte heterogeneity remains largely elusive, heterogeneity in radial glia across brain regions and within midbrain has been reported (La Manno et al. 2016; Li et al. 2023). Furthermore, different molecular and physiological features and distinct responses to stimuli were also observed in astrocytes from different mouse brain regions (Takata and Hirase 2008; Chai et al. 2017; Morel et al. 2017; Itoh et al. 2018; Kostuk et al. 2019; Makarava et al. 2019; Xin et al. 2019; Lozzi et al. 2020). Indeed, astrocyte heterogeneity has been suggested to underly the regional susceptibility to human diseases (Schober et al. 2022). Therefore, recapitulating astrocyte regional specificity in PSC-derived astrocytes is generally accepted as an important prerequisite.

Several reports have described the generation of regional astrocytes from human embryonic stem cells or induced pluripotent stem cells (iPSCs) which include forebrain (Krencik et al. 2011; Zhou et al. 2016; Tcw et al. 2017; Lin et al. 2018; Bradley et al. 2019; Hedegaard et al. 2020; Peteri et al. 2021), ventral midbrain (Booth et al. 2019; Barbuti et al. 2020; Crompton et al. 2021; de Rus Jacquet et al. 2021), hindbrain and spinal cord (Roybon et al. 2013; Serio et al. 2013; Holmqvist et al. 2015; Bradley et al. 2019; di Domenico et al. 2019; Yun et al. 2019). The regional identity of these astrocytes is typically evaluated at the stage of early neural progenitors generated via cell type- or region-directed neural patterning protocols, with the assumption that the positional characteristics will be faithfully preserved in the final astrocyte products. However, astrocyte production in vitro involves an extended period of astrocytic fate induction and progenitor expansion using FGF and EGF while substantial literature reported alterations in region-specific gene expression and/or neurogenic competence in expanded neural progenitors (Jain et al. 2003; Sun et al. 2008; Koch et al. 2009; Falk et al. 2012). Therefore, better characterisations of PSC-derived astrocytes and their lineage-specific features are needed to advance our knowledge about the molecular heterogeneity of human astrocytes.

Using a human iPSC line that allows the tracing of LMX1A expressing midbrain floor plate neural progenitors and their differentiated progeny (Cardo et al. 2023), we discovered an unexpected gradual depletion of LMX1A+ progenitor progeny during astrocyte induction from a bulk population of ventral midbrain patterned progenitors despite LMX1A+ progenitors being the predominant starting population. LMX1A+ progenitor derived astrocytes can however be generated if astrocytic induction is initiated from purified LMX1A+ progenitors, indicating that the positional constituents of the founding cell population may not be preserved faithfully in the derived astrocytes. Single cell RNA sequencing (scRNAseq) of astrocytes derived from both parental populations identified distinct transcriptomic signatures, providing a useful resource for assessment of human PSC-derived midbrain astrocytes.

Results

Depletion of LMX1A+ progenitors and/or derivatives in ventral midbrain patterned neural progenitor cultures during astrogenic induction

To investigate whether regionally patterned neural progenitors retain their lineage identity during astrogenic induction and glial progenitor expansion, we made use of the LMX1A-Cre/AAVS1-BFP iPSCs tracer line, which enables the tracking of LMX1A+ midbrain floor plate progenitors and their differentiated progeny(Cardo et al. 2023). We differentiated the LMX1A-Cre/AAVS1-BFP iPSCs towards the ventral midbrain fate following a modified protocol based on Jaeger et al. and Nolbrant et al.(Jaeger et al. 2011; Nolbrant et al. 2017) (Figure 1A). Immunocytochemistry of day (d)19 cultures confirmed a high proportion of cells expressing BFP (96.01±0.42%) and ventral midbrain progenitor markers LMX1A (92.94±0.91%), FOXA2 (94.76±0.57%) and OTX2 (97.82±0.28%; Figure 1B-D with the original images shown in Figure S1A). Most cells (91.26±1.64%) co-expressed LMX1A and FOXA2 (Figure 1B and 1D). At this stage, all BFP+ cells also stained positive to a pan neural progenitor marker NESTIN (Figure 1C). We detected a small proportion of cells expressing midbrain basal plate marker NKX6.1 (2.03±0.47%, Figure 1B-C) whose expression domain in the early developing ventral midbrain partially overlap with that of (Andersson et al. 2006)(Andersson et al. 2006). However, few PAX6+ cells were present (Figure 1C), which marks the forebrain and (Duan et al. 2013)(Duan et al. 2013). Immunocytochemical analysis of FACS-sorted BFP+ cells confirmed highly enriched expression of LMX1A (95.51±0.09%) and FOXA2 (95.35±0.14%), and co-expression of both markers (94.68±0.10%; Figure 1B and 1D). In contrast, only a small number of LMX1A+ cells (4.96±0.70%) were present in the sorted BFP- population (Figure 1B and1D). These findings provide further support that BFP expression faithfully recapitulate LMX1A expression and that LMX1A+ ventral midbrain progenitors represent the major cell population in d19(Cardo et al. 2023)(Cardo et al. 2023).

Depletion of LMX1A+ progenitors and their derivatives during astrogenic induction in ventral midbrain patterned neural progenitor cultures.

A, Schematic diagram of ventral midbrain neural differentiation and astrogenic induction. B-C, Representative view of immunocytochemistry of ventral midbrain neural progenitor markers and other regional markers in d19 unsorted, sorted BFP+, and sorted BFP- population. Scale bar represents 100 µm. Images shown were cropped to 300 µm×300 µm by randomly selecting the region of interest in the nuclei-only channel (uncropped greyscale images are shown in Figure S1A). D, Quantification of marker expression in unsorted, sorted BFP+, and sorted BFP- population. Error bars represent the standard error of means (SEM) of three independent experiments. E, Flow cytometry quantification of unsorted and BFP+ population during astrogenic induction and progenitor expansion. Each data point represents one biological replicate. The gating strategy used is shown in Figure S1B.

The d19 cells were then induced to undergo astrogenic switch in media containing FGF2 and EGF with the BFP content monitored by flow cytometry at each weekly passaging (Figure 1A, representative gating strategy is shown in Figure S1B-E). We found, unexpectedly, a dramatic decrease of BFP+ cell proportion from the starting point of 88.20±1.10% to only 27.59±8.28% at d43 and to nearly absent as the differentiation continued (Figure 1E). We did not observe evident cell death during culture and replating; thus, the absence of BFP could be either due to the silencing of BFP expression in the derivatives of LMX1A+ progenitors or the loss of these cells through growth competition. To address this question, we performed progenitor expansion and astrogenic induction under the same culture condition with purified d19 BFP+ progenitors isolated using fluorescence-activated cell sorting (FACS). Interestingly, the sorted BFP+ cells exhibited similar population growth rate to that of unsorted cultures, and the proportion of BFP+ cells remained at approximately 90% throughout the astrogenic induction and glial progenitor expansion period (Figure 1E). This observation demonstrates that BFP expression can be maintained in the derivatives of LMX1A+ midbrain floorplate progenitors, and that the loss of BFP+ cells in the unsorted culture is likely due to their growth disadvantage compared to the derivatives of LMX1A- progenitors.

Our finding is unexpected and demonstrates that the regional or lineage identity of PSC-derived astrocytic cells should not be assumed merely based on the dominant regional property of their cellular origin, given that no in vitro fate specification paradigm is 100% efficient.

Astrogenic switch occurred earlier in derivatives of LMX1A+ midbrain progenitors

Since sorted BFP+ (LMX1A+ or their derivatives) and unsorted astrocytic cultures differ distinctively in BFP expression soon after the initiation of astrogenic induction, for simplicity, these cultures are hereafter referred to as the BFP+ and BFP- cultures, respectively. Wondering whether the two cell populations behave differently in the process of astrogenic switch, we examined astrocytic marker expression in these cultures at d45 and d98 by immunocytochemistry. SOX9 and NFIA are transcription factors crucial for the initiation of astrogenesis and acquisition of astrogenic competence in the developing central nervous system (Stolt et al. 2003; Deneen et al. 2006), while CD44 identifies astrocyte-restricted precursors (Liu et al. 2004). We found that all these markers were more abundantly detected in the BFP+ cultures (NFIA: 65.89±2.81%; SOX9: 57.19±4.25%) than the BFP- cultures (NFIA: 4.26±1.28%; SOX9: 8.88±1.82%) at d45 (Two-way ANOVA with post-hoc Turkey test, NFIA: p=2.52×10-5, SOX9: p=9.21×10-6; Figure 2A-B with the original images shown in Figure S2). Although the number of NFIA+ and SOX9+ cells significantly increased in the BFP- cultures by d98 (NFIA: 44.07±4.56% on d98, p=3.58×10-4; SOX9: 44.28±2.84% on d98, p=1.21×10-4), the BFP+ cultures still contained more cells expressing NFIA (65.71±4.25%; p=2.06×10-2) and SOX9 (73.25±2.12%; p=5.51×10-4) than BFP-cultures (Figure 2A and 2C).

Early astrogenic switch and astrocyte maturation in derivatives of LMX1A+ midbrain progenitors.

A, Representative view of immunocytochemistry of astrogenic marker expression in BFP+ and unsorted progenitors at day 45 and 98. Scale bar represents 100 µm. Images shown were cropped to 462 µm×462 µm by randomly selecting the region of interest in the nuclei-only channel (uncropped greyscale images are shown in Figure S2). B, Representative view of immunocytochemistry of astrocyte marker expression in early and late, BFP+ and BFP- (unsorted) astrocytes. Scale bar represents 100 µm. Images shown were cropped to 300 µm×300 µm by randomly selecting the region of interest in the nuclei-only channel (uncropped greyscale images are shown in Figure S3). C, Quantification of immunocytochemistry of astrogenic marker in BFP+ and BFP- progenitors at day 45 and 98 shown in Panel A. Error bars represent the standard error of means (SEM) of three independent experiments. Two-way ANOVA was performed to compare between lineages (NFIA: p=5.389×10-6, df=1, effect size=3.62; SOX9: p=1.96×10-6, df=1, effect size=4.77) and days of differentiation (NFIA: p=7.82×10-5, df=1, effect size=1.99; SOX9: p=2.62×10-5, df=1, effect size=2.99) D, Quantification of immunocytochemistry of astrocyte marker expression in astrocytes. Error bars represent SEM of three independent experiments. Kruskal-Wallis test results following Bonferroni correction are shown on the top of the figure (AQP4: p.adjust=0.12, df=3, H=8.95; EAAT2: p.adjust=1.00, df=3, H=0.95; GFAP: p.adjust=0.06, df=3, H=10.38; S100B: p.adjust=0.11, df=3, H=9.05). E, Averaged trace of ATP-induced Ca2+ response assayed using FLIPR. Drugs or DMSO were applied at 1 minute of the assay. The line represents the average fluorescence change (ΔF/F0) in at least three independent experiments each with at least three replicate wells. The shaded area represents the SEM across at least three independent experiments. F, Quantitative comparison of the peak amplitude of ATP-induced Ca2+ response among conditions (two-way ANOVA, p<2.2×10-16, df=2, effect size=2.54) and samples (p=2.87×10-14, df=3, effect size=2.17). Error bars represent the SEM across at least independent experiments. G: Quantitative comparison of the rise time of ATP-induced Ca2+ response among conditions (two-way ANOVA, p=2.19×10-13, df=2, effect size =1.958) and samples (p=0.064, df=3, effect size=0.76). Intergroup comparison was performed using Post-hoc Tukey test. Error bars represent the SEM across at least three independent experiments. (****: p<0.0001, ***: p<0.001, **: p<0.01, *: p<0.05, ns: not significant).

To investigate whether the temporal difference in astrocytic switch between the BFP+ and BFP- cultures affects maturation and functionality of the derived astrocytes, we initiated astrocyte terminal differentiation by exposing the BFP+ and BFP- astrocyte precursors to CNTF and BMP4 (Krencik et al. 2011; Bradley et al. 2019) from d87 (referred to as early astrocytes) and d136 (late astrocytes). Both BFP+ and BFP-cultures exhibited a similar expression profile of classic astrocyte markers, including AQP4, EAAT2, and S100B, but few GFAP+ cells at both time points (Figure 2B and 2D with the original images shown in Figure S3).

As a reference, we also generated neural progenitors without employing any patterning cues and induced astrogenic switch and astrocyte differentiation from these non-patterned neural progenitors (Figure S4A-B). We found that, while the astrocyte cultures derived from the non-patterned progenitors contained a similar proportion of cells expressing AQP4, EAAT2 and S100B compared to the BFP+ and BFP- astrocyte cultures, there are more GFAP+ cells in the non-patterned astrocyte preparations (17.89±5.4%, Figure S4A-B).

Functional astrocytes exhibit transient calcium (Ca2+) spikes upon chemical stimulation, such as ATP (Zhang et al. 2016). Using a FLIPR Ca2+ release assay, we observed a sharp increase in the intracellular Ca2+ concentration upon ATP administration in both the early and late BFP+ astrocyte populations (early BFP+: p=5.37×10-12; late BFP+: p<2.2×10-16; Figure 2E-F). ATP induced Ca2+ release is partially mediated by inositol trisphosphate. Indeed, addition of an inositol trisphosphate receptor antagonist 2-aminoethoxydiphenylborate (2-APB) reduced the amplitude (early BFP+: p=5.88×10-5; late BFP+: p=9.22×10-10; Figure 2F) and rise time (early BFP+: p=2.63×10-5; late BFP+: p=3.59×10-4; late BFP-: p=0.028; Figure 2G) of ATP-induced Ca2+ response in both the early and late BFP+ astrocytes. Interestingly, the early BFP+ astrocytes had significantly lower peak amplitude than that observed in late BFP+ astrocytes (p=6.92×10-9; Figure 2F), despite their similar level of astrocyte marker expression, suggesting a difference in maturity at the functional level. Early and late BFP- astrocytes exhibited a similar profile of time-dependent increase in the amplitude of ATP-induced Ca2+ response but did not reach statistical significance (Figure 2E-F). However, late BFP- astrocytes showed a significantly lower peak amplitude than late BFP+ astrocytes (p<2.2×10-16; Figure 2E-F).

Taken together, our data demonstrate that astrogenesis occurred earlier in BFP+ cultures than BFP- cells. This temporal difference is also reflected in functional maturity of the derived astrocytes, despite a similar expression profile of classic astroglial markers.

Single cell RNA sequencing confirms the authenticity of PSC-derived astrocytes

To further characterize the PSC-derived astrocytes, we performed full-length scRNAseq on early and late BFP+ and BFP- astrocytes using the iCELL8 platform and SMART-seq technology, with non-patterned astrocytes derived from the LMX1A-Cre/AAVS1-BFP tracer line as a control. A sample of iPSC-derived neurons was included to facilitate downstream cell type identification. We profiled FACS purified astrocytes expressing CD49f as well as unsorted cultures for comparison of all three astrocyte populations (Barbar et al. 2020); Figure S4C-H). After stringent filtering (Figure S5A-C, see Methods and Materials on filtering), we obtained a total of 17478 protein-coding genes in 1786 qualifying cells, with an average of 6326 protein-coding genes detected per cell.

Unsupervised Louvain clustering identified 12 cell clusters (Figure 3A). Cells clustered mainly based on sample type (astrocytes and neurons; Figure S6A) and the estimated cell cycle phase (Figure S6B), while sorted CD49f+ and unsorted astrocytes were clustered together (Figure S6A). TagBFP was detected at a higher level in BFP+ astrocyte samples than in BFP- and NP samples, while TagBFP expression was negligible in the neuronal samples derived from an iPSC line without BFP transgene (Figure S6C-D). Using a set of known astrocyte and neuronal signature genes (Figure 3B), we identified cells in clusters 0, 1, and 5-11 as astrocytes (Figure 3B), which were enriched in expression of SOX9, NFIA, NFIB, HES1, S100A13, S100A16, EGFR, CD44 and GJA1 (Figure 3B). These transcripts were also detected at high levels in clusters 2 and 4, which were mostly estimated to be in cell cycle phases G2, M and S (Figure S6B). In addition, clusters 2 and 4 showed high levels of proliferation-related transcripts, such as TOP2A, MKI67, CDK1 and AURKA (Figure 3B), and are thus defined as astrocyte precursors. In contrast, Cluster 3 contains mostly cells from the neuronal sample (Figure 3A and Figure S6A) and indeed expressed high levels of genes closely related to neuronal structure and function (such as STMN2, SYT1, DCX, MAPT, and SNAP25; Figure 3B). We did not detect transcripts indicative of endoderm (GATA4), mesoderm (TBXT and TBX6), and oligodendrocyte progenitors (SOX10 and PDGFRA) in any of these clusters (Figure S6C).

Single cell RNA sequencing confirms the authenticity of PSC-derived astrocytes.

A, Uniform manifold approximation and projection plot of unbiased clustering, coloured by clusters. B, Heatmap of the normalised expression of selected markers in different clusters. The assigned identity to each cluster is shown at the top of the plot. C, Sankey plot summarising the result of reference mapping of cells in different clusters to eight published reference human brain scRNAseq datasets. The thickness of the thread is proportional to the number of cells mapped to the same identity in the reference datasets (predicted identity). Detailed results of referencing mapping to each reference datasets are shown in Figure S7A-H and prediction score shown in Figure S7I. (PRE: precursors; N: neurons)

To determine the authenticity of these PSC-derived astrocytes, we mapped our data to published scRNAseq datasets obtained from five foetal, an adult human brain and a PSC-derived astrocyte study using Seurat integration (La Manno et al. 2016; Sloan et al. 2017; Zhong et al. 2018; Polioudakis et al. 2019; Agarwal et al. 2020; Fan et al. 2020; Bhaduri et al. 2021; Eze et al. 2021). We found that cells in clusters annotated as astrocytes (clusters 0, 1, and 5-11) were indeed predominantly mapped to the reference astrocyte or astrocyte precursor populations with high confidence (prediction score over 0.5; Figure 3C and Figure S7). In contrast, neuronal cluster 2 was mapped to neurons of the foetal reference datasets, while the astrocyte precursor clusters (2 and 4) were mapped to progenitor populations in the foetal reference datasets (Figure 3C and Figure S7). These findings demonstrate that our iPSC-derived astrocytes closely resemble those in the human brains.

Distinct transcriptome fingerprints of LMX1A+ midbrain floor plate-derived astrocytes

Significant advance has been made recently in understanding the molecular profiles of midbrain dopamine neurons. However, our knowledge about midbrain astrocytes in this regard remains limited and does not inform anatomic or lineage origin of the cells. In this regard, the BFP+ astrocytes provide a unique resource to determine the transcriptomic characteristics of human midbrain floor plate derived astrocytes. By performing pairwise differential gene expression (details described in Methods and Materials), we identified 1153 genes differentially expressed (DEGs; adjusted p values less than 0.05 and log2 fold change over 0.25) in BFP+ astrocytes when compared to either BFP- or non-patterned astrocyte populations (Supplementary Data 1). Of these, 159 were unique to BFP+ astrocytes (BFP+ enriched, Figure 4A), which include genes associated with midbrain dopamine neuron development such as SULF1, LMO3, NELL2, and RCAN2 (Figure 4B) (Strelau et al. 2000; La Manno et al. 2016; Bifsha et al. 2017; Ahmed et al. 2021). Interestingly, LMX1A and FOXA2, which was used to evaluate PSC-derived midbrain astrocytes in previous studies, were not detected in BFP+ astrocytes (Figure S6D). We also identified 530 DEGs enriched in BFP- astrocytes only (BFP- enriched, Figure 4A and Supplementary Data 1), which include those known to be expressed in the ventrolateral - dorsal domain of the midbrain and hindbrain, such as IRX3, IRX5, PAX3, and PAX7 (Figure 4B)(Houweling et al. 2001; Matsunaga et al. 2001). This transcription profile supports the notion that the BFP- astrocytes are descendants of the initial minor populations of lateral midbrain progenitors. Moreover, 72 DEGs were shared by BFP+ and BFP- astrocytes against the non-patterned astrocytes (midbrain enriched, Figure 4A and Supplementary Data 1). This set of genes includes NR2F1, NR2F2, ZEB2, KCNJ6 and SRPX (Figure 4B), which have been reported to be signatures of mouse midbrain astrocytes (Endo et al. 2022). Together, our findings provide a new entry to transcriptomic characteristics of midbrain astrocytes and specifically a gene expression map of midbrain floor plate derived human astrocyte lineage.

Distinct transcriptome fingerprints of LMX1A+ midbrain floor plate-derived astrocytes

A, Heatmap of the normalised expression of population-specific genes in different populations of astrocytes. B, Violin plots of the normalised expression of selected candidate markers for BFP+, BFP-, and non-patterned (NP) astrocytes. C-D, Representative GO terms significantly enriched in BFP+ (C) and BFP- (D) enriched genes. Semantically similar representative terms were shown with the same colour.

Gene ontology (GO) enrichment analysis was performed on the 1153 DEGs enriched in BFP+ astrocytes (Supplementary Data 2). The significantly enriched GO terms were mainly related to various aspects of metabolism, stress response, biosynthesis, lysosomal activity, and cellular respiration (Figure 4C and Supplementary Data 3). These biological processes have previously been shown to be disrupted by several mutations causing familial Parkinson’s disease (di Domenico et al. 2019; Barbuti et al. 2020; Sonninen et al. 2020). In contrast, the GO terms associated with 530 BFP- astrocyte enriched DEGs were mostly related to formation of extracellular matrix and tissue development (Figure 4D and Supplementary Data 2-3). The differential enrichment of GO terms implicates functional difference between the BFP+ and BFP- astrocytes and supports the need for generating regional specific astrocytes for disease modelling.

Discussion

Despite the general belief that recapitulating astrocyte lineage heterogeneity is necessary for stem cell-based disease modelling and cell transplantation, the extent of astrocyte heterogeneity in different brain regions and subregions remains largely elusive. By harnessing an LMX1A based lineage tracing human iPSC line and cutting edge scRNAseq technology, we show that astrocytes derived from the LMX1A+ midbrain floor plate progenitors, the same cells giving rise to midbrain dopaminergic neurons, possess distinct transcriptional landmarks from those derived from non-midbrain patterned neural progenitors as well as other midbrain patterned progenitors. Moreover, we discovered unexpected negative selection against derivatives of LMX1A+ progenitors during astrocyte induction and progenitor expansion. Our study highlights the need for careful characterisation of PSC-derived astrocytes and provides a transcriptomic fingerprint for midbrain floor plate derived astrocytes.

Using a popular astrocyte in vitro differentiation paradigm, we found that astrocytes descended from the LMX1A+ midbrain progenitors could only be obtained from purified progenitors. In contrast, astrocytes derived from bulk midbrain patterned progenitors exhibits transcriptomic profiles of the lateral-dorsal midbrain, despite LMX1A+ midbrain progenitors being the predominant starting population. Our findings demonstrate that the lineage composition of the parent progenitors may not be faithfully preserved during astrocyte induction and progenitor expansion. FGF is the most popular inductive molecule used for astrocyte differentiation from stem cells (Chandrasekaran et al. 2016). It is evident however that FGF expanded neural progenitors, originated either from the brain or neutralized PSCs, exhibit restricted regional competence and positional gene expression. For example, bulk expanded human ventral midbrain neural progenitors (Jain et al. 2003), and fetal forebrain or spinal cord derived neural stem (NS) cells only give rise to GABAergic neurons (Sun et al. 2008); likewise, lt-NES cells display anterior hindbrain-like positional profile (Falk et al. 2012), while their antecedents, PSC-derived neural rosettes and early passage derivatives, express anterior forebrain markers (Koch et al. 2009). It is not clear whether this is due to deregulation of the original patterning at the level of gene expression or the loss of associated cell population (Gabay et al. 2003). In this study, since BFP+ astrocytes can be generated under the same culture condition with purified LMX1A+ progenitors, we reasoned that the loss of their derivatives in unsorted cultures was possibly due to differential growth capacity.

Our study highlights the need for careful assessment of astrocyte positional identity. A common practice in this regard is to confirm the regional characteristics of the founder progenitors following fate directed neural induction, with the assumption that the dominant positional features will be maintained by the astrocyte progeny (Krencik et al. 2011). This strategy is at least partly dictated by our limited knowledge in gene expression signatures of regional- and/or lineage-specific astrocytes. Hence, an end point evaluation of PSC-derived astrocytes often relies on region-specific markers defined in the developmental brain during the neurogenic period. For example, LMX1A and FOXA2 expression were used as criteria for midbrain astrocytes in previous studies (Barbuti et al. 2020; Crompton et al. 2023). However, scRNAseq of human fetal ventral midbrain and adult substantia nigra revealed negligible expression of these transcripts in astrocytes (La Manno et al. 2016; Agarwal et al. 2020; Kamath et al. 2022). Consistent with these findings, we did not detect LMX1A or FOXA2 in neither the BFP+ nor BFP- astrocytes. Our analysis however identified new positive and negative markers that could be applied to confirm the regional identity of ventral midbrain astrocytes.

In addition to distinct transcriptomic profile, BFP+ and BFP- astrocytes may also be functionally different. Astrocytes generated from progenitors broadly patterned to the dorsal forebrain, ventral forebrain and spinal cord have been shown to exhibiting different GO enrichment profile as well as different physiological and functional properties (Bradley et al. 2019). In comparison to the BFP- and non-patterned astrocytes, the current study revealed that GO terms enriched in BFP+ astrocytes, which originated from the same progenitor giving rise to midbrain dopaminergic neurons, were closely related to various biological processes disrupted in astrocytes carrying familial Parkinson’s disease mutations (di Domenico et al. 2019; Barbuti et al. 2020; Sonninen et al. 2020). Such a distinct enrichment profile implicates BFP+ astrocytes being functionally adapted to supporting midbrain dopaminergic neurons compared to BFP- and non-patterned astrocytes.

In conclusion, this study lends further support on regional diversity of astrocytes and identified a set of midbrain enriched genes. Crucially, the transcriptomic fingerprint of human midbrain floor plate-derived astrocytes described here offers a much-needed resource for assessing the authenticity of stem cell derived astrocytes in studies associated with Parkinson’s disease.

Methods and Materials

Stem cell culture and astrocyte differentiation

KOLF2 human iPSCs were maintained in E8 flex media (ThermoFisher) and manually dissociated using Gentle Cell Dissociation Reagent (STEMCELL Technologies) as previously described (Cardo et al. 2023). Astrocytes were differentiated using a three-stage stepwise strategy consisting neural induction and regional patterning, astrogenic switch and progenitor expansion, and astrocyte terminal differentiation. Midbrain floor plate progenitors were generated as previously described (Cardo et al. 2023). At day 19, cells were replated as single cells onto poly-D-lysine-laminin-coated plates at 1×106 cells/cm2 for astrogenic switch and progenitor expansion in N2B27 media supplemented with 10 ng/mL FGF2 (Peprotech) and 10 ng/mL Human EGF (Peprotech) and replated every 6-8 days. For astrocyte terminal differentiation, expanded neural progenitors were re-plated at a density of 3×104 cells/cm2 in expansion media and 24 hours later switched to N2B27 supplemented with 10 ng/mL human recombinant CNTF (Peprotech) and 10 ng/mL human recombinant BMP4 (Peprotech) for 7 days followed by media containing CNTF alone for another 13 days. 10 µM Y-27632 was used for 24 hours before and after each replating. The protocol for generating non-patterned astrocytes was the same as for floor plate-derived astrocyte except the neural progenitors were derived with duo-SMAD inhibitors only without ventral patterning reagents.

Flow cytometry analysis and cell isolation

Cells were dissociated in Accutase as described above and washed twice with DPBS by centrifugation for 5 minutes at 200 rcf. For evaluating BFP expression, dissociated cells were resuspended in 0.5 mM EDTA in DPBS (Sigma-Aldrich) and analysed on a BD LSRFortessa cell analyser (BD Biosciences). For purifying BFP+ cells, dissociated cells were resuspended in the same cell culture media. Background autofluorescence was compensated for using KOLF2 parental cell line at a similar stage of differentiation to define BFP- gating. For purifying CD49f+ astrocytes, dissociated cells were stained with Alexa Fluor 647-conjugated rat anti-CD49f antibody (5% v/v in a 100 µL reaction; BD Biosciences) for 25 minutes at 37°C on an orbital shaker at 200 rcf and resuspended in DPBS containing 0.5% bovine serum albumin and 50 units/mL DNase I (Sigma Aldrich). Background autofluorescence was compensated for using KOLF2 parental cell line at a similar stage of differentiation to define BFP- gating and unstained astrocytes to define CD49f- gating. Cell sorting was performed on a BD FACSAria III (BD Biosciences) using an 80 µm nozzle. Sorted cells were collected in the same media as for resuspension. Flow cytometry data were analysed in FlowJo v10.8.1 (BD Biosciences) as shown in Figure S1B-E. Briefly, non-debris events were selected using the eclipse gates on dot graphs of SSC-A versus FSC-A. Singlet events were sequentially gated using polygonal gates on dot graphs of FSC-H versus FSC-A and SSC-H versus SSC-A by selecting the events in the diagonal region. The positive and negative gates in the fluorescence channel were set as bifurcate gates at a minimum of 99.9% percentile (usually at 99.99% percentile) on the histogram of the fluorescence intensity of the negative control sample of the same flow cytometry experiment and applied to all samples of the same flow cytometry experiment.

Immunocytochemistry

Cultures were fixed with 3.7% PFA for 15-20 min at 4 °C. For nuclear antigen detection, an additional fixation with methanol gradient was performed, which include 5 mins each in 33% and 66% methanol at room temperature followed by 100% methanol for 20 min at -20°C. Cultures were then returned to PBST via inverse gradient and were then permeabilized with three 10-minute washes in 0.3% Triton-X-100 in PBS (PBS-T) and then blocked in PBS-T containing 1% BSA and 3% donkey serum. Cells were incubated with primary antibodies in blocking solution overnight at 4°C. Following three PBS-T washes, Alexa-Fluor secondary antibodies (Thermo Fisher Scientific) were added at 1:1000 PBS-T for 1 hour at ambient temperature in the dark. Three PBS-T washes were then performed that included once with DAPI (Molecular Probes). Images were taken on a Leica DMI6000B inverted microscope. Quantification was carried out in Cell Profiler (Stirling et al. 2021) or manually using ImageJ (Schindelin et al. 2012) by examining at least four randomly selected fields from three independent experiments. The antibodies used are provided in the Supplementary Table 1. Representative images shown in main figures were cropped by randomly selecting the region of interest in the DAPI-stained channel only, with the original unedited images shown in Figure S1A, S2 and S3.

Single-cell RNA-sequencing

Cells were dissociated with Accutase with 10 units/mL of papain (Sigma-Aldrich) for 10 minutes at 37°C and resuspended in 0.5% bovine serum albumin (Sigma Aldrich) with 50 units/mL DNase I in DPBS without calcium or magnesium (Gibco) and passed through a cell strainer with 35 µm mesh. Cells were stained with 1 µM SYTO16 (Invitrogen) and 0.08% (v/v) propidium iodide (part of the Invitrogen ReadyProbes™ Cell Viability Imaging Kit, Blue/Red) for 20 minutes on ice and then dispensed into the nano-well plates of the ICELL8® cx Single-Cell System (Takara). Wells containing single viable cells were automatically selected using the ICELL8 cx CellSelect v2.5 Software (Takara) with the green and not red logic. Manual triage was performed to recover additional candidate wells that contain viable single cells. The library for sequencing was then prepared using the SMART-Seq ICELL8 application kit (Takara) following the manufacturer’s recommended protocol. Next-generation sequencing was performed using the NovaSeq6000 and the Xp Workflow on a S4 flow cell for 200 cycles of pair-end sequencing.

Single-cell RNA-sequencing analysis

Using the Cogent NGS Analysis Pipeline software v 1.5.1 (Takara), FASTQ files containing all indices for each chip were demultiplexed to FASTQ files containing one index per file. Adaptor sequences were removed using cutadapt 3.2 with the following settings: -m 15 --trim-n --max-n 0.7 -q 20. Trimmed FASTQ files were aligned to the Homo sapiens GRCh38.106 primary assembly with the BFP reporter gene attached to the end of the genome, using STAR 2.7.9a (Dobin et al. 2013) with the following settings: --outSAMtype BAM Unsorted --quantMode TranscriptomeSAM -- outReadsUnmapped Fastx --outSAMstrandField intronMotif --chimSegmentMin 12 -- chimJunctionOverhangMin 8 --chimOutJunctionFormat 1 --alignSJDBoverhangMin 10 --alignMatesGapMax 100000 --alignIntronMax 100000 --alignSJstitchMismatchNmax 5 -1 5 5 --chimMultimapScoreRange 3 --chimScoreJunctionNonGTAG -4 -- chimMultimapNmax 20 --chimNonchimScoreDropMin 10 --peOverlapNbasesMin 12 - -peOverlapMMp 0.1 –alignInsertionFlush Right -- alignSplicedMateMapLminOverLmate 0 --alignSplicedMateMapLmin 30. After alignment, gene-level quantification was performed using featureCounts from subread 2.0.0 (Liao et al. 2013) with the following settings: -t exon --primary -R CORE -F GTF -Q 0 -B -g gene_id. The count matrix of each index was combined in R 4.2.0 (Team 2023).

All downstream analysis was performed in R 4.3.0 using Seurat 4.3.0 (Stuart et al. 2019). Gene-level filtering was applied by including only protein-coding genes with at least five total counts across all cells and being expressed in at least 1% of all cells. Poor quality cells were then identified using the is.outlier function from the scater 1.28.0 (McCarthy et al. 2017). Poor quality cells were defined as having a high percentage of mitochondrial gene count, or high or low the total number of genes detected, or high total gene counts. The thresholds of each metrics for each sample were determined as twice the median absolute deviation of the sample. Raw gene counts were log normalised with a scale.factor setting of 1×105. Data from the two batches of experiments were integrated using the FindIntegrationAnchors and IntegrateData based on the common top 2000 highly variable genes and the first 30 dimensions of principal components. The percentage of mitochondrial gene count and total gene count were regressed out using the ScaleData function. Principal component analysis (PCA) was performed on the top 2000 high variable genes and the number of principal components used for uniform manifold approximation and projection (UMAP) was determined using the JackStraw method (Chung and Storey 2015). UMAP and unbiased Louvain clustering was performed on the first 33 principal components. Pairwise differential gene expression analysis was performed using the MAST method (Finak et al. 2015) with “Chip” as the latent variable. Gene ontology enrichment analysis was performed using enrichGO function in the clusterProfiler 4.10.0 package (Stirling et al. 2021) with all genes in the filtered dataset as the background. GO term database was downloaded using the org.Hs.eg.db 3.18.0 package (Carlson 2019). Revigo v1.8.1 was used to group the representative GO terms based on semantic similarity using a size setting of 0.5, Homo sapiens database, and SimRel method for semantic similarity (Schlicker et al. 2006; Supek et al. 2011).

Published datasets were downloaded from NCBI’s Gene Expression Omnibus (Clough and Barrett 2016) and processed in R 4.2.0. Gene level filtering was performed by retaining only protein-coding genes with more than five total counts across all cells. Gene counts were normalised using the NormalizeData function (scale.factor settings are listed in Supplementary Table 2). PCA was performed based on the top 2000 highly variable genes to obtain the first 50 PCs. Visual inspection of the elbow plot was used to determine the number of PCs for downstream analysis. Batch effect between subjects was evaluated on the two-dimensional PC2∼PC1 plot. Where inter-subject batch effect was observed, Harmony integration was performed based on the PCs selected in the previous step (Supplementary Table 2). UMAP was performed based on either the PCA or Harmony reduction (using the top 30 dimensions), and Louvain clustering was performed (settings shown in Supplementary Table 2). Cluster identities were verified against the reported annotation where possible. For datasets without detailed annotation published or astrocyte lineage reported (Supplementary Table 2), reannotation was performed based on the expression of known markers (Figure S7A-E). Reference mapping was performed using FindTransferAnchors and TransferData function in Seurat based on the first 30 dimensions of either the PCA or Harmony loadings of the reference dataset.

Statistical analyses

All data were collected from at least three independent experiments and presented as mean ± standard error of means unless otherwise specified. Data were tested for normality with the Shapiro-Wilk test and for equal variance with Levene test before performing statistical analyses by two-way ANOVA with post-hoc Tukey test for multiple comparisons where relevant. Kruskal-Wallis test with post-hoc Dunn’s test for pairwise comparison was used where parametric test was not suitable. Effect size was calculated as Cohen’s f for ANOVA or eta squared based on the H-statistic for Kruskal-Wallis test. All statistical tests were performed in R4.3.0.

Supporting information

eLife Assessment

The manuscript by Li and coworkers analyzed astrocytic differentiation of midbrain floor plate-patterned neural cells originating from human iPS cells, with a LMX1A reporter. This valuable work identifies transcriptomic differences at the single-cell level, between astrocytes generated from LMX1A reporter positive or negative cells, as well as non-patterned astrocytes and neurons. The evidence is solid, but the paper can be strengthened by further analyses of the transcriptomic data, and astrocytic morphology; also, searching for some of the differentially expressed genes by immunohistochemistry in different regions of the mammalian brain, or in human specimens, would be very informative.

Reviewer #1 (Public Review):

Summary:

In a previous study, the authors developed a human iPS cell line which expresses Cre under the control of the Lmx1a promoter in order to track, select for, and differentiate human dopamine neurons. In the manuscript under review, the authors are using methods which they have modified to generate astrocytes from the same cell line. The authors are interested in examining astrocytes which are derived from regionalized, floor plate progenitors.

The fundamental weakness of this paper is that the authors are making arguments about regional identity but their work is limited to experiments in vitro. Some of the claims that the authors make should be tested in vivo - ie, in sections, at least. Are floor plate markers or other ventral markers ever expressed in astrocytes or glial progenitors in the mammalian fetus? When do astrocytes emerge in the floor plate? All of the data here are based on an overly simplified in vitro platform.

Lmx1a expression is not limited to the ventral midbrain; it is also expressed in other parts of the developing, ventral CNS and in the roof plate and dorsal CNS (Millonig et al, Nature 2000). Indeed, many of the phenotypes of the Lmx1a mutant mouse (dreher) have little to do with the ventral midbrain. The authors are making an assumption that regional identity is fixed when they begin their astrocyte differentiation protocol - not necessarily true. After astrocytic differentiation is initiated, the authors have done little to demonstrate that floor plate identity is maintained even in selected cells; in fact, the transcriptomic data suggests that the cells are released from a floor plate fate. The authors seem to realize this but do not make any attempt to prove their thesis. If regional identity is not maintained, the authors need a better experiment.

If regional identity is not maintained, so what? Don't we already know that this can happen? The authors acknowledge that this is known in the discussion.

The authors have done transcriptomics studies to follow the changes in these cells but they have not told us very much that is meaningful. It would be useful to validate some of the new astrocytic markers that they have identified - Pax and Irx genes (Welle et al., Glia 2021) come quickly to mind. What about genes related to Shh and Wnt signaling that are prevalent in the floor plate? In particular, a lot of work has been done examining the role of Shh on the properties and lineage of astrocytes (Farmer et al., Science 2016; Hill et al., eLife 2019; Gingrich et al., Neural Dev 2022; Xie et al., Cell Rep 2022). There are a lot of stones which remain unturned, here, and the authors could actually tell us much more without doing an immense amount of work. These suggestions and criticisms are described in far greater detail in the confidential comments to the authors.

Work Cited:

Chizhikov et al., Mamm Genome 2006. https://pubmed.ncbi.nlm.nih.gov/17019651/

Chizhikov et al., Development 2004. https://journals.biologists.com/dev/article/131/11/2693/42269/Control-of-roof-plate-formation-by-Lmx1a-in-the

Chizhikov et al., PNAS 2010. https://pubmed.ncbi.nlm.nih.gov/20498066/

Emsley and Macklis. Neuron Glia Biol 2007. https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1820889/

Farmer et al., Science 2016. https://pubmed.ncbi.nlm.nih.gov/26912893/

Gross et al., Development 2016. https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4958331/

Hill et al., eLife 2019. https://pubmed.ncbi.nlm.nih.gov/31194676/

Gingrich et al., Neural Dev 2022. https://pubmed.ncbi.nlm.nih.gov/35027088/

Iskusnykh et al., eLife 2023. https://elifesciences.org/articles/84095

Millonig et al, Nature 2000. https://pubmed.ncbi.nlm.nih.gov/10693804/

Welle et al. Glia 2021. https://pubmed.ncbi.nlm.nih.gov/36342840/

Xie et al., Cell Rep 2022. https://pubmed.ncbi.nlm.nih.gov/35196485/

Reviewer #2 (Public Review):

In the current manuscript Li et al., study the preservation of the regional identity during the process of astrocyte generation from pluripotent stem cells. More precisely, this work investigates if neural progenitor cells patterned for the ventral midbrain give rise to astrocytes with conserved regional specification, which could reflect the astrocytic heterogeneity in the brain. To this end, the authors utilized a previously generated reporter iPSC line in which the expression of introduced blue fluorescence protein (BFP) is subjacent to the activation of LMXA1, a ventral midbrain floor plate marker. The study reports that following a defined patterning protocol based on SHH and FGF8, over 90% of d19 cells, corresponding to a neural progenitor stage, acquired the midbrain floor plate identity. However, during the subsequent astrogenic induction and glial progenitor expansion, this identity is gradually lost, supposedly due to the growth advantage of cells deriving from the residual LMX1A- neural progenitors. Contrariwise, if the LMX1A+ progenitors were purified, regional identity would be maintained throughout the astrocytic generation and incur an early astrogenic switch and maturation of derived astrocytes. By using single-cell RNA sequencing, the authors further identified distinct transcriptomic signatures on the astrocytic progeny of LMX1A- and LMX1A- progenitors.

Strengths and weaknesses:

(1) The main model utilized was engineered from the KOLF2 human iPSC line into an elegant LMX1A-reporter line based on the expression of BFP. This results in an attractive model for studies tracing the fate of LMX1A cells. However, consideration should be given to the fact that the parental line, exhibits a splice disruption in the COL3A1 gene encoding type III collagen (Pantazis 2022, doi:10.1016/j.stem.2022.11.004 ), which has been identified as being enriched in certain ventral astrocytic populations (Bradley 2019, doi:10.1242/dev.170910).

(2) The authors argue that the depletion of BFP seen in the unsorted population immediately after the onset of astrogenic induction is due to the growth advantage of the derivatives of the residual LMX1A- population. However, no objective data supporting this idea is provided, and one could also hypothesize that the residual LMX1A- cells could affect the overall LMX1A expression in the culture through negative paracrine regulation. Therefore, cell cycle or proliferation studies of these cells are needed to prove the authors' assumption. Furthermore, on line 124 it is stated that: "Interestingly, the sorted BFP+ cells exhibited similar population growth rate to that of unsorted cultures...". In the face of the suggested growth disadvantage of those cells, this statement needs clarification.

(3) Regarding the fidelity of the model system, it is not clear to me how the TagBFP expression was detected in the BFP+ population supposedly in d87 and d136 pooled astrocytes (Fig S6C) while no LMX1A expression was observed in the same cells (Fig S6F).

(4) The generated single-cell RNASeq dataset is extremely valuable. However, given the number of conditions included in this study (i.e. early vs late astrocytes, BFP+ vs BFP-, sorted vs unsorted, plus non-patterned and neuronal samples) the resulting analysis lacks detail. For instance, from a developmental perspective and to better grasp the functional significance of astrocytic heterogeneity, it would be interesting to map the identified clusters to early vs late populations and to the BFP status. Moreover, although comprehensive, Figure S7 is complex to understand given that citations rather than the reference populations are depicted.

(5) Do the authors have any consideration regarding the morphology of the astrocytes obtained in this study? None of the late astrocyte images depict a prototypical stellate morphology, which is reported in many other studies involving the generation of iPSC-derived astrocytes and which is associated with the maturity status of the cell.

Author response:

The following is the authors’ response to the original reviews

Reviewer 1:

In vitro nature of this work being the fundamental weakness of this paper

We disagree with this statement. As explained in the provisional response, the aim of this study was to test the validity of a general concept applied in pluripotent stem cell research that pluripotent stem cell-derived astrocytes faithfully represent the lineage heterogeneity of their ancestral neural progenitors and hence preserve the regionality of such progenitors. Our genetic lineage study is justified for addressing this in vitro-driven question. However, we have highlighted the rationale where appropriate in the revised paper.

If regional identity is not maintained, so what? Don't we already know that this can happen? The authors acknowledge that this is known in the discussion.

Importance of regional identity: Growing evidence demonstrates the functional heterogeneity of brain astrocytes in health and disease. Therefore, for in vitro disease modeling, it is believed that one should use astrocytes represent the anatomy of disease pathology; for example, midbrain astrocytes for studying dopamine neurodegeneration and Parkinson’s disease. Understanding the dynamics of stem cell-derived astrocytes and identifying astrocyte subtypes is important for their biomedical applications.

Regional identity change/Discussion: It seems that the reviewer misunderstood the context in which the ‘identity change’ was discussed. The literature referred to (in the Discussion) concerns shifts in regional gene expression in bulk-cultured cells. In the days of pre-single-cell analysis/lineage tracking, one cannot distinguish whether this was due to a change in the transcriptomic landscape in progenies of the same lineage or alterations in lineage heterogeneity, but to interpret at face value as regional identity was not maintained. In the revised paper, we have made an effort to indicate that ‘regional identity’ is used broadly to refer to lineage relationships and/or traits rather than static gene expressioin.

validation of the markers/additional work

The scNAseq analysis performed in this study compared the profiles of astrocytes derived from LMX1A+ and LMX1A- ventral midbrain-patterned neural progenitors. Since it is not possible to perform genetic lineage tracking in humans and an analogous mouse lineage tracer line is not available, in vivo validation of these markers with respect to their lineage relationship is not currently feasible. However, we took advantage of abundant single-cell human astrocyte transcriptomic datasets and validated our genes in silico. We also validated the differential expression of selected markers in late BFP+ and BFP- astrocytes using immunocytochemistry, where reliable antibodies are available. The results of the additional analyses are presented in Figure S8 and Supplemental Data 5.

Knowledge gaps concerning astrocyte development

Reviewer 1 pointed out a number of knowledge gaps concerning astrocyte development, such as the transcriptomic landscape trajectories of midbrain floor plate cells as they progress towards astrocytes. Indeed, the limited knowledge on regional astrocyte molecule heterogeneity restricts the objective validation of in vitro-derived astrocyte subtypes and the development of novel approaches for their generation in vitro. We agree with the need for in-depth in vivo studies using model organisms, although these are beyond the scope of the current work.

Reviewer 2:

(1) The authors argue that the depletion of BFP seen in the unsorted population immediately after the onset of astrogenic induction is due to the growth advantage of the derivatives of the residual LMX1A- population. However, no objective data supporting this idea is provided, and one could also hypothesize that the residual LMX1A- cells could affect the overall LMX1A expression in the culture through negative paracrine regulation.

We acknowledge the lack of evidence-based explanation for the depletion of BFP+ cells in mixed cultures. We were unable to perform additional experiments because of resource limitations. The design of the LMX1A-Cre/AAVS1-BFP lineage tracer line determines that BFP is expressed irreversibly in LMX1A-expressing cells or their derivatives regardless of their LMX1A expression status. Therefore, the potential negative paracrine regulation of LMX1A by residual LMX1A- cells should not affect cells that have already turned on BFP. We have highlighted the working principles of the LMX1A tracer line in the revised manuscript.

(2) Furthermore, on line 124 it is stated that: "Interestingly, the sorted BFP+ cells exhibited similar population growth rate to that of unsorted cultures...". In the face of the suggested growth disadvantage of those cells, this statement needs clarification.

To avoid confusion, we have removed the statement.

(3) Regarding the fidelity of the model system, it is not clear to me how the TagBFP expression was detected in the BFP+ population supposedly in d87 and d136 pooled astrocytes (Fig S6C) while no LMX1A expression was observed in the same cells (Fig S6F).

The TagBFP tracer is expressed in the progenies of LMX1A+ cells, regardless of their LMX1A expression status. We have gone through the MS text to ensure that this information has been provided.

(4) The generated single-cell RNASeq dataset is extremely valuable. However, given the number of conditions included in this study (i.e. early vs late astrocytes, BFP+ vs BFP-, sorted vs unsorted, plus non-patterned and neuronal samples) the resulting analysis lacks detail. For instance, from a developmental perspective and to better grasp the functional significance of astrocytic heterogeneity, it would be interesting to map the identified clusters to early vs late populations and to the BFP status.

We performed additional bioinformatics analysis, which provided independent support for the relative developmental maturity suggested by functional assays. The additional data are now provided in the revised Figure 3B, C, E.

Moreover, although comprehensive, Figure S7 is complex to understand given that citations rather than the reference populations are depicted.

The information provided in the revised Figure S7.

(5) Do the authors have any consideration regarding the morphology of the astrocytes obtained in this study? None of the late astrocyte images depict a prototypical stellate morphology, which is reported in many other studies involving the generation of iPSC-derived astrocytes and which is associated with the maturity status of the cell.

The morphology of our astrocytes was not unique to the present study. Many factors may influence the morphology of astrocytes, such as the culture media and supplements used, and maturity status. Based on the functional assays and limited GFAP expression, our astrocytes were relatively immature.

Version 2

Abstract

Abstract

The regional specificity of stem cell-derived astrocytes is believed to be an important prerequisite for their application in disease modelling and cell-based therapies. Due to the lack of subtype-defining markers for astrocytes in different regions of the brain, the regional identity of in vitro-derived astrocytes is often declared by the dominant positional characteristics of their antecedent neural progenitors, patterned to a fate of interest, with the assumption that the positional trait is preserved by the derived astrocytes via linear descent. Using a human induced pluripotent stem cell line designed for tracing derivatives of LMX1A+ cells combined with a ventral midbrain induction paradigm, we show that astrocytes originating from LMX1A+ progenitors can only be generated if these progenitors are purified prior to the astrocyte differentiation process, or their progenies are gradually lost to progenies of LMX1A- progenitors. This finding indicates that the lineage composition of iPSC-derived astrocytes may not accurately recapitulate the founder progenitor population. Using deep single-cell RNA sequencing, we identified distinct transcriptomic signatures in astrocytes derived from the LMX1A+ progenitor cells. Our study highlights the need for rigorous characterization of pluripotent stem cell-derived regional astrocytes, and provides a resource for assessing LMX1A+ ventral midbrain progenitor-derived human astrocytes.

Introduction

Astrocytes are the most abundant cell type in the brain. They play important roles in the central nervous system in supporting neuronal survival and synaptic activities, including regulation of ionic homeostasis, providing energetic support, elimination of oxidative stress, and neurotransmitter removal and recycling (Verkhratsky and Nedergaard 2018). Abnormalities in astrocytes have been linked to various neurodegenerative and neurodevelopmental disorders such as Parkinson’s disease, Alzheimer’s disease, Huntington’s disease, autism spectrum disorders, and Alexander’s disease (Molofsky et al. 2012; Phatnani and Maniatis 2015; Booth et al. 2017). Therefore, there is a growing interest in using human pluripotent stem cell (PSC)-derived astrocytes for in vitro disease modeling (Chandrasekaran et al. 2016).

Contrary to the widely held belief that astrocytes in the brain are largely identical, recent studies have revealed a diversity in their transcriptomic profiles, physiological properties, and functions (Oberheim et al. 2009; Schober et al. 2022). Single-cell and spatial transcriptomic studies have identified several astrocyte subpopulations in the mouse cortex (Zhu et al. 2018; Batiuk et al. 2020; Bayraktar et al. 2020). In humans, although astrocyte heterogeneity remains largely elusive, heterogeneity in radial glia across brain regions and within the midbrain has been reported (La Manno et al. 2016; Li et al. 2023). Furthermore, different molecular and physiological features as well as distinct responses to stimuli have been observed in astrocytes from different mouse brain regions (Takata and Hirase 2008; Chai et al. 2017; Morel et al. 2017; Itoh et al. 2018; Kostuk et al. 2019; Makarava et al. 2019; Xin et al. 2019; Lozzi et al. 2020). Indeed, astrocyte heterogeneity has been suggested to underlie the regional susceptibility to human diseases (Schober et al. 2022). Therefore, recapitulating astrocyte regional specificity in PSC-derived astrocytes is generally accepted as an important prerequisite.

Several studies have described the generation of regional astrocytes from human embryonic stem cells or induced pluripotent stem cells (iPSCs), including the forebrain (Krencik et al. 2011; Zhou et al. 2016; Tcw et al. 2017; Lin et al. 2018; Bradley et al. 2019, Hedegaard et al. 2020; Peteri et al. 2021), and ventral midbrain (Booth et al. 2019; Barbuti et al. 2020; Crompton et al. 2021; de Rus Jacquet et al. 2021), hindbrain and spinal cord (Roybon et al. 2013; Serio et al. 2013; Holmqvist et al. 2015; Bradley et al. 2019; di Domenico et al. 2019; Yun et al. 2019). The regional identity of these astrocytes is typically evaluated at the stage of early neural progenitors generated via cell type- or region-directed neural patterning protocols, with the assumption that the positional characteristics will be faithfully preserved in the final astrocyte products. However, astrocyte production in vitro involves an extended period of astrocytic fate induction and progenitor expansion using FGF and EGF, and substantial literature has reported alterations in region-specific gene expression and/or neurogenic competence in expanded neural progenitors (Jain et al. 2003; Sun et al. 2008; Koch et al. 2009; Falk et al. 2012). Therefore, better characterization of PSC-derived astrocytes and their lineage-specific features is needed to advance our knowledge of the molecular heterogeneity of human astrocytes.

Using a human iPSC line that allows the tracing of LMX1A expressing ventral midbrain neural progenitors and their differentiated progeny (Cardo et al. 2023), we discovered an unexpected gradual depletion of LMX1A+ progenitor progeny during astrocyte induction from a bulk population of ventral midbrain patterned progenitors, despite LMX1A+ progenitors being the predominant starting population. However, LMX1A+ progenitor-derived astrocytes can be generated if astrocytic induction is initiated from purified LMX1A+ progenitors, indicating that the positional constituents of the founding cell population may not be preserved faithfully in the derived astrocytes. Single-cell RNA sequencing (scRNAseq) of astrocytes derived from both parental populations identified distinct transcriptomic signatures, providing a useful resource for the assessment of a defined lineage of human PSC-derived midbrain astrocytes.

Results

Depletion of LMX1A+ progenitors and/or derivatives in ventral midbrain patterned neural progenitor cultures during astrogenic induction

To investigate whether regionally patterned neural progenitors retain their lineage identity during astrogenic induction and glial progenitor expansion, we made use of the LMX1A-Cre/AAVS1-BFP iPSCs tracer line, in which LMX1A-driven Cre activates BFP expression, thereby enabling tracking of LMX1A+ ventral midbrain progenitors and their differentiated progeny (Cardo et al. 2023). We differentiated the LMX1A-Cre/AAVS1-BFP iPSCs towards the ventral midbrain fate following a modified protocol based on Jaeger et al. and Kriks et al (Jaeger et al. 2011; Kriks et al. 2011; Figure 1A). Immunocytochemistry of day (d) 19 cultures confirmed a high proportion of cells expressing BFP (96.01±0.42%) and ventral midbrain progenitor markers LMX1A (92.94±0.91%), FOXA2 (94.76±0.57%), and OTX2 (97.82±0.28%; Figure 1B-D with the original images shown in Figure S1A). Most cells (91.26±1.64%) co-expressed LMX1A and FOXA2 (Figure 1B and 1D). At this stage, all BFP+ cells stained positive for the pan-neural progenitor marker NESTIN (Figure 1C). We detected a small proportion of cells expressing the midbrain basal plate marker NKX6.1 (2.03±0.47%, Figure 1B-C), whose expression domain in the early developing ventral midbrain partially overlap with that of LMX1A (Andersson et al. 2006). However, few PAX6+ cells were present (Figure 1C), which marks the forebrain and lateral midbrain (Duan et al. 2013). Immunocytochemical analysis of FACS-sorted BFP+ cells confirmed highly enriched expression of LMX1A (95.51±0.09%) and FOXA2 (95.35±0.14%), and co-expression of both markers (94.68±0.10%; Figure 1B and 1D). In contrast, only a small number of LMX1A+ cells (4.96±0.70%) were present in the sorted BFP- population (Figure 1B and1D). These findings provide further support that BFP expression faithfully recapitulates LMX1A expression, and that LMX1A+ ventral midbrain progenitors represent the major cell population at d19 (Cardo et al. 2023).

Depletion of LMX1A+ progenitors and their derivatives during astrogenic induction in ventral midbrain patterned neural progenitor cultures.

A, Schematic diagram of ventral midbrain neural differentiation and astrogenic induction. B-C, Representative view of immunocytochemistry of ventral midbrain neural progenitor markers and other regional markers in d19 unsorted, sorted BFP+, and sorted BFP- population. Scale bar represents 100 µm. Images shown were cropped to 300 µm × 300 µm by randomly selecting the region of interest in the nuclei-only channel (uncropped greyscale images are shown in Figure S1A). D, Quantification of marker expression in unsorted, sorted BFP+, and sorted BFP- population. Error bars represent the standard error of means (SEM) of three independent experiments. E, Flow cytometry quantification of unsorted and BFP+ population during astrogenic induction and progenitor expansion. Each data point represents one biological replicate. The gating strategy used is shown in Figure S1B.

The d19 cells were then induced to undergo astrogenic switch in media containing FGF2 and EGF with the BFP+ cells (LMX1A+ ventral midbrain progenitors or their derivatives) monitored by flow cytometry at each weekly passaging (Figure 1A, representative gating strategy is shown in Figure S1B-E). Unexpectedly, we found a dramatic decrease in the BFP+ cell proportion from the starting point of 88.20±1.10% to only 27.59±8.28% at d43 and nearly absent as differentiation continued (Figure 1E). We did not observe any evident cell death during culture and replating; thus, the absence of BFP could be either due to the silencing of BFP expression in the derivatives of LMX1A+ progenitors or the loss of these cells through growth competition. To address this question, we performed progenitor expansion and astrogenic induction under the same culture conditions as purified d19 BFP+ progenitors isolated using fluorescence-activated cell sorting (FACS). Interestingly, the proportion of BFP+ cells remained at approximately 90% throughout the astrogenic induction and glial progenitor expansion periods (Figure 1E). This observation demonstrates that BFP expression can be maintained in the derivatives of LMX1A+ midbrain progenitors and that the loss of BFP+ cells in the unsorted culture is likely due to their growth disadvantage compared to the derivatives of LMX1A- progenitors. Our findings are unexpected and demonstrate that the regional or lineage identity of PSC-derived astrocytic cells should not be assumed based merely on the dominant lineage identity of their cellular origin, given that no in vitro fate specification paradigm is 100% efficient.

Astrogenic switch occurred earlier in derivatives of LMX1A+ midbrain progenitors

Since sorted BFP+ (LMX1A+ ventral midbrain progenitors or their derivatives) and unsorted astrocytic cultures differ distinctively in BFP expression soon after the initiation of astrogenic induction, for simplicity, these cultures are hereafter referred to as the BFP+ and BFP- cultures, respectively. To determine whether the two cell populations behave differently in the process of astrogenic switch, we examined astrocytic marker expression in these cultures at d45 and d98 using immunocytochemistry. SOX9 and NFIA are transcription factors that are crucial for the initiation of astrogenesis and acquisition of astrogenic competence in the developing central nervous system (Stolt et al. 2003; Deneen et al. 2006), whereas CD44 has been identified as an astrocyte-restricted precursor (Liu et al. 2004). We found that all these markers were more abundantly detected in the BFP+ cultures (NFIA: 65.89±2.81%; SOX9:57.19±4.25%) than in the FP cultures (NFIA: 4.26±1.28%; SOX9:8.88±1.82%) at d45 (Two-way ANOVA with post-hoc Tukey test, NFIA: p=2.52×10-5, SOX9: p=9.21×10-6; Figure 2A and 2C with the original images shown in Figure S2). Although the number of NFIA+ and SOX9+ cells significantly increased in the BFP- cultures by d98 (NFIA: 44.07±4.56% on d98, p=3.58×10-4; SOX9:44.28±2.84% on d98, p=1.21×10-4), the BFP+ cultures still contained more cells expressing NFIA (65.71±4.25%; p=2.06×10-2) and SOX9 (73.25±2.12%; p=5.51×10-4) than in BFP- cultures (Figure 2A and 2C).

Early astrogenic switch and astrocyte maturation in derivatives of LMX1A+ midbrain progenitors.

A, Representative view of immunocytochemistry of astrogenic marker expression in BFP+ and unsorted progenitors at day 45 and 98. Scale bar represents 100 µm. Images shown were cropped to 462 µm × 462 µm by randomly selecting the region of interest in the nuclei-only channel (uncropped greyscale images are shown in Figure S2). B, Representative view of immunocytochemistry of astrocyte marker expression in early and late, BFP+ and BFP- (unsorted) astrocytes. Scale bar represents 100 µm. Images shown were cropped to 300 µm × 300 µm by randomly selecting the region of interest in the nuclei-only channel (uncropped greyscale images are shown in Figure S3). C, Quantification of immunocytochemistry of astrogenic marker in BFP+ and BFP- progenitors at day 45 and 98 shown in Panel A. Error bars represent the standard error of means (SEM) of three independent experiments. Two-way ANOVA was performed to compare between lineages (NFIA: p=5.389×10-6, df=1, effect size=3.62; SOX9: p=1.96×10-6, df=1, effect size=4.77) and days of differentiation (NFIA: p=7.82×10-5, df=1, effect size=1.99; SOX9: p=2.62×105, df=1, effect size=2.99) D, Quantification of immunocytochemistry of astrocyte marker expression in astrocytes. Error bars represent SEM of three independent experiments. Kruskal-Wallis test results following Bonferroni correction are shown on the top of the figure (AQP4: p.adjust=0.12, df=3, H=8.95; EAAT2: p.adjust=1.00, df=3, H=0.95; GFAP: p.adjust=0.06, df=3, H=10.38; S100B: p.adjust=0.11, df=3, H=9.05). E, Averaged trace of ATP-induced Ca2+ response assayed using FLIPR. Drugs or DMSO were applied at 1 minute of the assay. The line represents the average fluorescence change (ΔF/F0) in at least three independent experiments each with at least three replicateΔwells. The shaded area represents the SEM across at least three independent experiments. F, Quantitative comparison of the peak amplitude of ATP-induced Ca2+ response among conditions (two-way ANOVA, p<2.2×10-16, df=2, effect size=2.54) and samples (p=2.87×10-14, df=3, effect size=2.17). Error bars represent the SEM across at least independent experiments. G: Quantitative comparison of the rise time of ATP-induced Ca2+ response among conditions (two-way ANOVA, p=2.19×10-13, df=2, effect size =1.958) and samples (p=0.064, df=3, effect size=0.76). Intergroup comparison was performed using post-hoc Tukey test. Error bars represent the SEM across at least three independent experiments. (****: p<0.0001, ***: p<0.001, **: p<0.01, *: p<0.05, ns: not significant).

To investigate whether the temporal difference in the astrocytic switch between the BFP+ and BFP- cultures affects the maturation and functionality of the derived astrocytes, we initiated astrocyte terminal differentiation by exposing the BFP+ and BFP- astrocyte precursors to CNTF and BMP4 (Krencik et al. 2011; Bradley et al. 2019) from d87 (referred to as early astrocytes) and d136 (late astrocytes). Both BFP+ and BFP- cultures exhibited a similar expression profile of classic astrocyte markers, including AQP4, EAAT2, and S100B, but few GFAP+ cells at both time points (Figure 2B and 2D with the original images shown in Figure S3). As a reference, we also generated neural progenitors without employing any patterning cues and induced astrogenic switch and astrocyte differentiation from these non-patterned neural progenitors (Figure S4A-B). We found that, while the astrocyte cultures derived from the non-patterned progenitors contained a similar proportion of cells expressing AQP4, EAAT2 and S100B compared to the BFP+ and BFP- astrocyte cultures, there are more GFAP+ cells in the non-patterned astrocyte preparations (17.89±5.4%, Figure S4A-B).

Functional astrocytes exhibit transient calcium (Ca2+) spikes upon chemical stimulation such as ATP (Zhang et al. 2016). Using a FLIPR Ca2+ release assay, we observed a sharp increase in the intracellular Ca2+ concentration upon ATP administration in both the early and late BFP+ astrocyte populations (early BFP+: p=5.37 10-12; late BFP+: p<2.2×10-16; Figure 2E-F). ATP-induced Ca2+ release is partially mediated by inositol trisphosphate. Indeed, addition of an inositol trisphosphate receptor antagonist 2-aminoethoxydiphenylborate (2-APB) reduced the amplitude (early BFP+: p=5.88×10-5; late BFP+: p=9.22×10-10; Figure 2F) and rise time (early BFP+: p=2.63×10-5; late BFP+: p=3.59×10-4; late BFP-: p=0.028; Figure 2G) of ATP-induced Ca2+ response in both the early and late BFP+ astrocytes. Interestingly, early BFP+ astrocytes had a significantly lower peak amplitude than that observed in late BFP+ astrocytes (p=6.92×10-9; Figure 2F), despite their similar levels of astrocyte marker expression, suggesting a difference in maturity at the functional level. Early and late BFP- astrocytes exhibited a similar profile of time-dependent increase in the amplitude of ATP-induced Ca2+ response, but did not reach statistical significance (Figure 2E-F). However, late BFP- astrocytes showed a significantly lower peak amplitude than late BFP+ astrocytes (p<2.2×10-16; Figure 2E-F).

Taken together, our data demonstrated that astrogenesis occurred earlier in BFP+ cultures than in BFP-cells. This temporal difference is also reflected in the functional maturity of the derived astrocytes despite a similar expression profile of classic astroglial markers.

Single cell RNA sequencing confirms the authenticity of PSC-derived astrocytes

To further characterize the PSC-derived astrocytes, we performed full-length scRNAseq on early and late BFP+ and BFP- astrocytes using the iCELL8 platform and SMART-seq technology, with non-patterned astrocytes derived from the LMX1A-Cre/AAVS1-BFP tracer line as a control. A sample of iPSC-derived neurons was included to facilitate the downstream cell type identification. We profiled FACS purified astrocytes expressing CD49f as well as unsorted cultures for comparison of all three astrocyte populations (Barbar et al. 2020); Figure S4C-H). After stringent filtering (Figure S5A-C, see Methods and Materials on filtering), we obtained 17478 protein-coding genes in 1786 qualifying cells, with an average of 6326 protein-coding genes detected per cell.

Unsupervised Louvain clustering identified 12 cell clusters (Figure 3A). Cells were clustered mainly based on sample type (astrocytes and neurons; Figure S6A) and the estimated cell cycle phase (Figure S6B), whereas sorted CD49f+ and unsorted astrocytes were largely intermingled together (Figure S6A). Clusters 0, 1, 4, 5 and 11 consisted of predominantly BFP+ astrocytes, each with varying contribution from the ‘early’ or ‘late’ astrocyte samples. Clusters 2, 6, and 7 were dominated by BFP-astrocytes, with the majority of the ‘early’ astrocyte samples in cluster 2. Cells in clusters 8,9 and 10 were primarily non-patterned astrocytes, whereas cluster 3 came from the neuronal sample (Figure 3B-C). Consistent with immunocytochemistry, TagBFP transcripts were detected at a higher level in BFP+ astrocyte samples than in BFP- and NP samples, while TagBFP expression was negligible in neuronal samples derived from an iPSC line without the BFP transgene (Figure S6C-D). Using a set of known astrocyte and neuronal signature genes (Figure 3D), we identified cells in clusters 0, 1, and 5-11 as astrocytes (Figure 3D), which were enriched in SOX9, NFIA, NFIB, HES1, S100A13, S100A16, EGFR, CD44, and GJA1 expression (Figure 3D). These transcripts were also detected at high levels in clusters 2 and 4, which were mostly estimated to be in the cell cycle phases G2, M, and S (Figure S6B). In addition, clusters 2 and 4 showed high levels of proliferation-related transcripts, such as TOP2A, MKI67, CDK1, and AURKA (Figure 3D) and were thus defined as astrocyte precursors. In contrast, Cluster 3 contained mostly cells from the neuronal sample (Figure 3A and Figure S6A) and expressed high levels of genes closely related to neuronal structure and function (such as STMN2, SYT1, DCX, MAPT, and SNAP25; Figure 3D). We did not detect transcripts indicative of endoderm (GATA4), mesoderm (TBXT and TBX6), or oligodendrocyte progenitors (SOX10 and PDGFRA) in any of these clusters (Figure S6C).

Single cell RNA sequencing confirms the authenticity of PSC-derived astrocytes.

A-B , Uniform manifold approximation and projection plot of unbiased clustering, coloured by clusters or sample group. C, Number of cells from different sample groups in each clusters. D, Heatmap of the normalised expression of selected markers in different clusters. The assigned identity to each cluster is shown at the top of the plot. E, Principal component projection analysis of pseudobulk astrocyte data onto a reference principal component axis of astrocyte maturity (Zhang et al. 2016; Hedegaard et al. 2020). Each dot represents a pseudobulk sample of one independent sample from each clusters. F, Sankey plot summarising the result of reference mapping of cells in different clusters to eight published reference human brain scRNAseq datasets. The thickness of the thread is proportional to the number of cells mapped to the same identity in the reference datasets (predicted identity). Detailed results of referencing mapping to each reference datasets are shown in Figure S7A-H and prediction score shown in Figure S7I. (PRE: precursors; N: neurons).

Next, we examined the developmental status of astrocytes by pseudobulk analysis using published bulk RNA-seq datasets from human fetal and postmortem astrocytes (Zhang et al. 2016). As expected, the astrocyte precursors (clusters 2 and 4) were projected to be less mature than the fetal astrocytes (Figure 3E). The BFP+ astrocyte dominant clusters (0, 1, 5, and 11) and non-patterned astrocyte clusters (8, 9, and 10) were shown to be more developmentally advanced than the BFP- astrocyte clusters (6 and 7). Thus, the pseudobulk analysis provides independent support to the functional assays suggesting the relative maturity of BFP+ and BFP- astrocytes (Figure 2).

To determine the authenticity of these PSC-derived astrocytes, we mapped our data to five published scRNAseq datasets obtained from human fetal and adult brains using Seurat integration (La Manno et al. 2016; Sloan et al. 2017; Zhong et al. 2018; Polioudakis et al. 2019; Agarwal et al. 2020; Fan et al. 2020; Bhaduri et al. 2021; Eze et al. 2021). We found that cells in clusters annotated as astrocytes (clusters 0, 1, and 5-11) were predominantly mapped to the reference astrocyte or astrocyte precursor populations with high confidence (prediction score > 0.5; Figure 3F and Figure S7). In contrast, neuronal cluster 3 was mapped to neurons in the fetal reference datasets, while the astrocyte precursor clusters (2 and 4) were mapped to progenitor populations in the fetal reference datasets (Figure 3F and Figure S7). These findings demonstrate that iPSC-derived astrocytes closely resemble those in the human brain.

Distinct transcriptome fingerprints of LMX1A+ midbrain progenitor-derived astrocytes

Significant advances have been made recently in the understanding of the molecular profiles of midbrain dopamine neurons. However, our knowledge of midbrain astrocytes in this regard remains limited and does not inform the anatomic or lineage origin of the cells. In this regard, BFP+ astrocytes provide a unique resource for determing the transcriptomic characteristics of human astrocytes derived from the LMX1A+ lineage of ventral midbrain patterned progenitors. By performing pairwise differential gene expression (details described in Methods and Materials), we identified 1153 genes differentially expressed (DEGs; adjusted p values less than 0.05 and log2 fold change over 0.25) in BFP+ astrocytes when compared to either BFP- or non-patterned astrocyte populations (Supplementary Data 1). Of these, 287 were unique to BFP+ astrocytes (BFP+ enriched, Figure 4A), including genes associated with midbrain dopamine neuron development such as SULF1, LMO3, NELL2, and RCAN2 (Figure 4B) (Strelau et al. 2000; La Manno et al. 2016; Bifsha et al. 2017; Ahmed et al. 2021). 55 of the 287 BFP+ DEGs, including SULF1, NELL2, RCAN2, and RGS5, were confirmed to be significantly preferentially expressed in midbrain astrocytes in vivo (adjusted p-values less than 0.05) by analysing the integration of five human brain datasets (Figure S8A-B; La Manno et al. 2016; Zhong et al. 2019; Fan et al. 2020; Eze et al. 2021). Another 87 BFP+ DEGs, such as SLC7A3, SIX6, and LMO3, were also detected at higher levels in midbrain astrocytes than in forebrain astrocytes, although their expression levels were not significant (Figure S8A-B). Moreover, LMX1A and FOXA2, which have been used to evaluate PSC-derived midbrain astrocytes in previous studies, were not detected in BFP+ astrocytes (Figure S6D, S8C).

Distinct transcriptome fingerprints of astrocytes derived from LMX1A+ ventral midbrain progenitors

A, Heatmap of the normalised expression of population-specific genes in different populations of astrocytes. B, Violin plots of the normalised expression of selected candidate markers for BFP+, BFP-, and non-patterned (NP) astrocytes. C-D, Representative GO terms significantly enriched in BFP+ (C) and BFP- (D) enriched genes. Semantically similar representative terms were shown with the same colour.

We also identified 159 DEGs enriched only in BFP- astrocytes (BFP- enriched, Figure 4A and Supplementary Data 1), including those known to be expressed in the ventrolateral-dorsal domain of the midbrain and hindbrain, such as IRX3, IRX5, PAX3, and PAX7 (Figure 4B) (Houweling et al. 2001; Matsunaga et al. 2001). Differential expression of PAX3 and PAX7 in BFP astrocytes was confirmed at the protein level by immunocytochemistry of d136 BFP+ and BFP- astrocyte cultures, whereas their expression in the human midbrain was validated by published datasets in silico (Figure S8). This transcription profile supports the notion that BFP astrocytes are descendants of initial minor populations of lateral midbrain progenitors. Moreover, 72 DEGs were shared by BFP+ and BFP- astrocytes compared to non-patterned astrocytes (midbrain enriched, Figure 4A and Supplementary Data 1). This set of genes included NR2F1, NR2F2, ZEB2, KCNJ6, and SRPX (Figure 4B), which have been reported to be signatures of mouse midbrain astrocytes (Endo et al. 2022). Together, our findings provide a new entry into the transcriptomic characteristics of midbrain astrocytes, specifically a gene expression map of the LMX1A+ midbrain progenitor-derived human astrocyte lineage.

Gene ontology (GO) enrichment analysis was performed on the 1153 DEGs enriched in BFP+ astrocytes (Supplementary Data 2). The significantly enriched GO terms were mainly related to various aspects of metabolism, stress response, biosynthesis, lysosomal activity, and cellular respiration (Figure 4C and Supplementary Data 3). These biological processes have previously been shown to be disrupted by several mutations that cause familial Parkinson’s disease (di Domenico et al. 2019; Barbuti et al. 2020; Sonninen et al. 2020). In contrast, the GO terms associated with 530 BFP- astrocyte enriched DEGs were mostly related to the formation of the extracellular matrix and tissue development (Figure 4D and Supplementary Data 2-3). The differential enrichment of GO terms implies functional differences between BFP+ and BFP- astrocytes and supports the need for generating regional specific astrocytes for disease modelling.

Discussion

Despite the general belief that recapitulating astrocyte lineage heterogeneity is necessary for stem cell-based disease modelling and cell transplantation, the extent of astrocyte heterogeneity in different brain regions, their anatomical origins, and associated molecular signatures remain largely elusive. This knowledge gap limits the end-point characterization of stem cell-derived astrocytes; hence, reliance on the dominant regional characteristics of the initial neural progenitor populations, with the assumption that the lineage representation of the progenitors is preserved in the derived astrocytes. By harnessing an LMX1A based lineage tracing human iPSC line, we discovered unexpected negative selection against derivatives of LMX1A+ midbrain-patterned progenitors during astrocyte induction and progenitor expansion, highlighting the need for careful characterization of PSC-derived astrocytes and reinforce the need for a deeper understanding of the molecular landscape of astrocytes in different regions of the human brain.

Most neural progenitors used for astrocyte differentiation in this study coexpressed LMX1A and FOXA2. In the ventral midbrain, the combinatorial expression of these transcription factors defines the dopaminergic neural progenitors (Failli et al. 2002; Andersson et al. 2006). We found that astrocytes derived from LMX1A+ progenitors could only be obtained if LMX1A+ cells were purified prior to astrocyte differentiation. In contrast, astrocytes derived from bulk midbrain-patterned progenitors exhibit transcriptomic profiles of the lateral-dorsal midbrain despite LMX1A+ progenitors being the predominant starting population. Our findings demonstrate that the lineage composition of parent progenitors was not preserved during astrocyte induction or progenitor expansion. FGF is the most commonly used inductive molecule for astrocyte differentiation from stem cells (Chandrasekaran et al. 2016). However, it is evident that FGF-expanded neural progenitors, originating either from the brain or neutralized PSCs, exhibit restricted regional competence and positional gene expression. For example, bulk-expanded human ventral midbrain neural progenitors (Jain et al. 2003), fetal forebrain or spinal cord derived neural stem (NS) cells only give rise to GABAergic neurons (Sun et al. 2008), and lt-NES cells display an anterior hindbrain-like positional profile (Falk et al. 2012), while their antecedents, PSC-derived neural rosettes and early passage derivatives, express anterior forebrain markers (Koch et al. 2009). It is unclear whether this is due to the deregulation of the original patterning at the level of gene expression or the loss of the associated cell population (Gabay et al. 2003). In this study, because BFP+ astrocytes can be generated under the same culture conditions as purified LMX1A+ progenitors, we reasoned that the loss of their derivatives in unsorted cultures was possibly due to their differential growth capacity.

Our study highlights the need for a careful assessment of the positional identity of in vitro-derived astrocytes. A common practice in this regard is to confirm the regional identity of founder progenitors following fate-directed neural induction, with the assumption that the dominant positional features are maintained by astrocyte progeny (Krencik et al. 2011). This strategy is, at least partly, dictated by our limited knowledge of the gene expression signatures of regional and/or lineage-specific astrocytes. Hence, endpoint evaluation of PSC-derived astrocytes often relies on region-specific markers defined in the brain during the neurogenic period. For example, LMX1A and FOXA2 expression has been used as criteria for midbrain astrocytes in previous studies (Barbuti et al. 2020; Crompton et al. 2023). However, scRNA-seq of the human fetal ventral midbrain and adult substantia nigra has revealed negligible expression of these transcripts in astrocytes (La Manno et al. 2016; Agarwal et al. 2020; Kamath et al. 2022). Consistent with these findings, we did not detect LMX1A or FOXA2 in BFP+ or BFP − astrocytes. However, our analysis identified new positive and negative markers that can be used to validate astrocytes derived from the LMX1A+ lineage of ventral midbrain progenitors. Future work will benefit from in vivo validation of these putative lineage-specific markers, as a mouse analogous tracer line is available, whereas lineage tracking is not possible in humans.

In addition to the distinct transcriptomic profiles, BFP+ and BFP astrocytes may also be functionally different. Astrocytes generated from progenitors broadly patterned to the dorsal forebrain, ventral forebrain, and spinal cord have been shown to exhibit different GO enrichment profiles as well as different physiological and functional properties (Bradley et al. 2019). In comparison to BFP- and non-patterned astrocytes, the current study revealed that GO terms enriched in BFP+ astrocytes, which originated from the same progenitor giving rise to midbrain dopaminergic neurons, were closely related to various biological processes disrupted in astrocytes carrying familial Parkinson’s disease mutations (di Domenico et al. 2019; Barbuti et al. 2020; Sonninen et al. 2020). Such a distinct enrichment profile suggests that BFP+ astrocytes may be functionally adapted to support midbrain dopaminergic neurons compared with BFP− and non-patterned astrocytes. Indeed, astrocytes isolated from the ventral midbrain have been reported to exhibit stronger neurotrophic effects and the ability to reduce α-synuclein aggregation in the midbrain than cortical astrocytes in cellular and mouse models of PD (Kostuk et al. 2019; Yang et al. 2022), highlighting the importance of understanding astrocyte heterogeneity in iPSC disease modelling.

In conclusion, this study provides further evidence for the regional diversity of astrocytes and identifies a set of midbrain-enriched genes. Crucially, the transcriptomic fingerprint of human astrocytes derived from LMX1A-expressing midbrain progenitors reported here offers a much-needed resource for assessing the authenticity of stem cell-derived astrocytes in studies of Parkinson’s disease.

Methods and Materials

Stem cell culture and astrocyte differentiation

KOLF2 human iPSCs were maintained in E8 flex media (ThermoFisher) and manually dissociated using Gentle Cell Dissociation Reagent (STEMCELL Technologies) as previously described (Cardo et al. 2023). Astrocytes were differentiated using a three-stage stepwise strategy consisting neural induction and regional patterning, astrogenic switch and progenitor expansion, and astrocyte terminal differentiation. LMX1A+ ventral midbrain progenitors were generated as previously described (Cardo et al. 2023). At day 19, cells were replated as single cells onto poly-D-lysine-laminin-coated plates at 1×106 cells/cm2 for astrogenic switch and progenitor expansion in N2B27 media supplemented with 10 ng/mL FGF2 (Peprotech) and 10 ng/mL Human EGF (Peprotech) and replated every 6-8 days. For astrocyte terminal differentiation, expanded neural progenitors were re-plated at a density of 3×104 cells/cm2 in expansion media and 24 hours later switched to N2B27 supplemented with 10 ng/mL human recombinant CNTF (Peprotech) and 10 ng/mL human recombinant BMP4 (Peprotech) for 7 days followed by media containing CNTF alone for another 13 days. 10 µM Y-27632 was used for 24 hours before and after each replating. The protocol for generating non-patterned astrocytes was the same as for LMX1A+ ventral midbrain-derived astrocyte except the neural progenitors were derived with duo-SMAD inhibitors only without ventral patterning reagents.

Flow cytometry analysis and cell isolation

Cells were dissociated in Accutase as described above and washed twice with DPBS by centrifugation for 5 minutes at 200 rcf. For evaluating BFP expression, dissociated cells were resuspended in 0.5 mM EDTA in DPBS (Sigma-Aldrich) and analysed on a BD LSRFortessa cell analyser (BD Biosciences). For purifying BFP+ cells, dissociated cells were resuspended in the same cell culture media.

Background autofluorescence was compensated for using KOLF2 parental cell line at a similar stage of differentiation to define BFP- gating. For purifying CD49f+ astrocytes, dissociated cells were stained with Alexa Fluor 647-conjugated rat anti-CD49f antibody (5% v/v in a 100 µL reaction; BD Biosciences) for 25 minutes at 37°C on an orbital shaker at 200 rcf and resuspended in DPBS containing 0.5% bovine serum albumin and 50 units/mL DNase I (Sigma Aldrich). Background autofluorescence was compensated for using KOLF2 parental cell line at a similar stage of differentiation to define BFP- gating and unstained astrocytes to define CD49f- gating. Cell sorting was performed on a BD FACSAria III (BD Biosciences) using an 80 µm nozzle. Sorted cells were collected in the same media as for resuspension. Flow cytometry data were analysed in FlowJo v10.8.1 (BD Biosciences) as shown in Figure S1B-E. Briefly, non-debris events were selected using the eclipse gates on dot graphs of SSC-A versus FSC-A. Singlet events were sequentially gated using polygonal gates on dot graphs of FSC-H versus FSC-A and SSC-H versus SSC-A by selecting the events in the diagonal region. The positive and negative gates in the fluorescence channel were set as bifurcate gates at a minimum of 99.9% percentile (usually at 99.99% percentile) on the histogram of the fluorescence intensity of the negative control sample of the same flow cytometry experiment and applied to all samples of the same flow cytometry experiment.

FLIPR calcium assay

Astrocytes were plated on day 10 of astrocyte terminal differentiation and cultured according to the standard protocol until day 20. On the day of recording, equal volume of FLIPR Calcium 6 Assay buffer was added to cells without replacing or washing and cells were incubated for 2 hours at 37°C. Drug assay buffers were prepared in HBSS with Ca2+ and Mg2+ (Gibco) with 20 mM HEPES buffer (Gibco) at 5X concentration. After incubation, cells were imaged on FLIPR Penta system (Molecular Devices) at a frequency of 2 Hz for 1 minute prior to injection. 25 µL of the drug assay buffers were dispensed at a speed of 25µL/second and images captured at a frequency of 2 Hz for a further 5 minutes. Raw fluorescence intensity data were exported and analysed in R 4.3.0 using a custom script to identify the peak response. The average baseline fluorescence intensity for each well was calculated by averaging the raw fluorescence intensity measured 60 seconds prior to drug application. The normalised change in fluorescence intensity above the baseline (ΔF/F0) for each well, at a given timepoint, was calculated using the following formula: where Ft is the fluorescence intensity at timepoint t, F0 is the average fluorescence intensity of the baseline period (60 seconds prior to drug application).

Immunocytochemistry

Cultures were fixed with 3.7% PFA for 15-20 min at 4 °C. For nuclear antigen detection, an additional fixation with methanol gradient was performed, which include 5 mins each in 33% and 66% methanol at room temperature followed by 100% methanol for 20 min at -20°C. Cultures were then returned to PBST via inverse 21 gradient and were then permeabilized with three 10-minute washes in 0.3% Triton-X-100 in PBS (PBS-T) and then blocked in PBS-T containing 1% BSA and 3% donkey serum. Cells were incubated with primary antibodies in blocking solution overnight at 4°C. Following three PBS-T washes, Alexa-Fluor secondary antibodies (Thermo Fisher Scientific) were added at 1:1000 PBS-T for 1 hour at ambient temperature in the dark. Three PBS-T washes were then performed that included once with DAPI (Molecular Probes). Images were taken on a Leica DMI6000B inverted microscope. Quantification was carried out in Cell Profiler (Stirling et al. 2021) or manually using ImageJ (Schindelin et al. 2012) by examining at least four randomly selected fields from three independent experiments. The antibodies used are provided in the Supplementary Table 1. Representative images shown in main figures were cropped by randomly selecting the region of interest in the DAPI-stained channel only, with the original unedited images shown in Figure S1A, S2 and S3.

Single-cell RNA-sequencing

Cells were dissociated with Accutase with 10 units/mL of papain (Sigma-Aldrich) for 10 minutes at 37°C and resuspended in 0.5% bovine serum albumin (Sigma Aldrich) with 50 units/mL DNase I in DPBS without calcium or magnesium (Gibco) and passed through a cell strainer with 35 µm mesh. Cells were stained with 1 µM SYTO16 (Invitrogen) and 0.08% (v/v) propidium iodide (part of the Invitrogen ReadyProbes™ Cell Viability Imaging Kit, Blue/Red) for 20 minutes on ice and then dispensed into the nano-well plates of the ICELL8® cx Single-Cell System (Takara). Wells containing single viable cells were automatically selected using the ICELL8 cx CellSelect v2.5 Software (Takara) with the green and not red logic. Manual triage was performed to recover additional candidate wells that contain viable single cells. The library for sequencing was then prepared using the SMART-Seq ICELL8 application kit (Takara) following the manufacturer’s recommended protocol. Next-generation sequencing was performed using the NovaSeq6000 and the Xp Workflow on a S4 flow cell for 200 cycles of pair-end sequencing.

Single-cell RNA-sequencing analysis

Using the Cogent NGS Analysis Pipeline software v 1.5.1 (Takara), FASTQ files containing all indices for each chip were demultiplexed to FASTQ files containing one index per file. Adaptor sequences were removed using cutadapt 3.2 with the 22 following settings: -m 15 --trim-n --max-n 0.7 -q 20. Trimmed FASTQ files were aligned to the Homo sapiens GRCh38.106 primary assembly with the BFP reporter gene attached to the end of the genome, using STAR 2.7.9a (Dobin et al. 2013) with the following settings: --outSAMtype BAM Unsorted --quantMode TranscriptomeSAM --outReadsUnmapped Fastx --outSAMstrandField intronMotif --chimSegmentMin 12 --chimJunctionOverhangMin 8 --chimOutJunctionFormat 1 –alignSJDBoverhangMin 10 --alignMatesGapMax 100000 --alignIntronMax 100000 -- alignSJstitchMismatchNmax 5 -1 5 5 --chimMultimapScoreRange 3 -- chimScoreJunctionNonGTAG -4 --chimMultimapNmax 20 -- chimNonchimScoreDropMin 10 --peOverlapNbasesMin 12 --peOverlapMMp 0.1 – alignInsertionFlush Right --alignSplicedMateMapLminOverLmate 0 -- alignSplicedMateMapLmin 30. After alignment, gene-level quantification was performed using featureCounts from subread 2.0.0 (Liao et al. 2013) with the following settings: -t exon --primary -R CORE -F GTF -Q 0 -B -g gene_id. The count matrix of each index was combined in R 4.2.0 (Team 2023).

All downstream analysis was performed in R 4.3.0 using Seurat 4.3.0 (Stuart et al. 2019). Gene-level filtering was applied by including only protein-coding genes with at least five total counts across all cells and being expressed in at least 1% of all cells. Poor quality cells were then identified using the is.outlier function from the scater 1.28.0 (McCarthy et al. 2017). Poor quality cells were defined as having a high percentage of mitochondrial gene count, or high or low the total number of genes detected, or high total gene counts. The thresholds of each metrics for each sample were determined as twice the median absolute deviation of the sample. Raw gene counts were log normalised with a scale.factor setting of 1×105. Data from the two batches of experiments were integrated using the FindIntegrationAnchors and IntegrateData based on the common top 2000 highly variable genes and the first 30 dimensions of principal components. The percentage of mitochondrial gene count and total gene count were regressed out using the ScaleData function. Principal component analysis (PCA) was performed on the top 2000 high variable genes and the number of principal components used for uniform manifold approximation and projection (UMAP) was determined using the JackStraw method (Chung and Storey 2015). UMAP and unbiased Louvain clustering was performed on the first 33 principal components. Pairwise differential gene expression analysis was performed using the MAST method (Finak et al. 2015) with “Chip” as the latent variable. Gene ontology enrichment analysis was performed using enrichGO function in the clusterProfiler 4.10.0 package (Stirling et al. 2021) with all genes in the filtered dataset as the background. GO term database was downloaded using the org.Hs.eg.db 3.18.0 package (Carlson 2019). Revigo v1.8.1 was used to group the representative GO terms based on semantic similarity using a size setting of 0.5, Homo sapiens database, and SimRel method for semantic similarity (Schlicker et al. 2006; Supek et al. 2011).

Published datasets were downloaded from NCBI’s Gene Expression Omnibus (Clough and Barrett 2016) and processed in R 4.2.0. Gene level filtering was performed by retaining only protein-coding genes with more than five total counts across all cells. Gene counts were normalised using the NormalizeData function (scale.factor settings are listed in Supplementary Table 2). PCA was performed based on the top 2000 highly variable genes to obtain the first 50 PCs. Visual inspection of the elbow plot was used to determine the number of PCs for downstream analysis. Batch effect between subjects was evaluated on the two-dimensional PC2∼PC1 plot. Where inter-subject batch effect was observed, Harmony integration was performed based on the PCs selected in the previous step (Supplementary Table 2). UMAP was performed based on either the PCA or Harmony reduction (using the top 30 dimensions), and Louvain clustering was performed (settings shown in Supplementary Table 2). Cluster identities were verified against the reported annotation where possible. For datasets without detailed annotation published or astrocyte lineage reported (Supplementary Table 2), reannotation was performed based on the expression of known markers (Figure S7A-E). Reference mapping was performed using FindTransferAnchors and TransferData function in Seurat based on the first 30 dimensions of either the PCA or Harmony loadings of the reference dataset.

Statistical analyses

All data were collected from at least three independent experiments and presented as mean ± standard error of means unless otherwise specified. Data were tested for normality with the Shapiro-Wilk test and for equal variance with Levene test before performing statistical analyses by two-way ANOVA with post-hoc Tukey test for multiple comparisons where relevant. Kruskal-Wallis test with post-hoc Dunn’s test for pairwise comparison was used where parametric test was not suitable. Effect size was calculated as Cohen’s f for ANOVA or eta squared based on the H-statistic for Kruskal-Wallis test. All statistical tests were performed in R4.3.0.